Exploring the Feasability of\\ Multifactor Authentication with EarEEG

### 2018-02-06 paper.tex
Round 1 of updates from previous rebuttal

\title{Towards One-Step, Three-Factor \\ Authentication with EarEEG}

@@

\begin{abstract}
Multifactor authentication presents a robust method to secure our private information, but typically requires multiple actions by the user resulting in a high cost to usability and limiting adoption. A usable system should also be unobtrusive and inconspicuous. We present a system that can provide all three factors of authentication (knowledge, possession, and inherence) in a single step via an earpiece which implements brain-based authentication using in-ear electroencephalography (EEG). We demonstrate its potential by collecting EEG data using manufactured custom-fit earpieces with embedded electrodes. Across all participants’ best-performing “passthoughts”, we are able to achieve 0\% false acceptance and 0.36\% false rejection rates, an overall accuracy of 99.82\%, using one earpiece with three electrodes. Furthermore, we find no successful attempts simulating impersonation attacks. We also report perspectives from our participants and discuss directions for future work. Our results indicate that a single electrode-laden earpiece could provide a discreet, convenient, and robust method for one-step, three-factor authentication.

@@

We find that we can achieve very low false acceptance rates (FAR) and false rejection rates (FRR) with a single, three-electrode earpiece. Interestingly, we found that performance improves in the ear compared to a single electrode on forehead above the left eye, a site typically used for collecting EEG data. Additionally, we find that passthoughts could not be spoofed by imposters, even when the imposters had the user's earpiece and knew their chosen passthoughts. Finally we discuss interesting patterns in our promising results, and future work to investigate the robustness of passthoughts in day-to-day life.

@@

These results thus far establish good performance in our default training strategy, in which we count as negative examples recordings from the wrong participant performing any task. For comparison, we report three other analyses with differing negative examples: the correct task recorded from the wrong participant (inherence only), the wrong task recorded from the correct participant (knowledge only), and a combination of these two. Positive examples were always the correct participant performing the correct task.

@@

Overall, our default training strategy achieves the lowest FAR (Table \ref{tab:compare}). Within-task FAR is ten times higher, and within-participants FAR is one hundred times higher, though for all scenarios still less than 1\%.
However, FRR is \textit{lower} in the within-tasks training strategy than the default. The highest FRR is in the combined negative example case, though FAR remains low.

Finally, our "leave one out" analysis with participants' best tasks maintained 0 FAR across all participant combinations.

@@

The first aspect of this scenario we tested was the ability of an imposter to wear an earpiece acquired from someone else and achieve viable impedance values for EEG collection based on the fit of the pieces in their ears. P1 tried on each of the other participants' customized earpieces, which were able to at least physically fit in P1's ears. The impedances of P1 wearing other participants' earpieces were then recorded and are listed in Table \ref{tab:imposter_impedances} below. While some are a better fit than others, overall they are higher (worse) than those achieved by the pieces' intended owners themselves.

\begin{table}[h]
\begin{center}
\begin{tabular}{lrrrrrr}
& \multicolumn{6}{c}{Impedance [k\(\Omega\)]} \\
\cline{2-7}
& \multicolumn{3}{|c|}{\textbf{Left ear}} & \multicolumn{3}{c|}{\textbf{Right ear}} \\
\textbf{P} & \textbf{C} & \textbf{F} & \textbf{B} & \textbf{C} & \textbf{F} & \textbf{B} \\
\hline
2 & 34.1 & 10.2 & 12.8 & 27.8 & 16.0 & 16.3\\
3 & 21.1 & 20.9 & 19.0 & 13.5 & 11.3 & 19.5\\
4 & 14.1 & 11.9 & 9.7 & 11.0 & 11.1 & 13.3\\
5 & 17.2 & 21.9 & 10.3 & 32.6 & 12.5 & 11.6\\
6 & 18.7 & 10.0 & 8.4 & 14.8 & 11.5 & 8.9\\
7 & 91.5 & \textgreater1000 & 21.5 & 33.5 & 26.4 & 31.0\\
\end{tabular}
\end{center}
\caption{Electrical impedances with P1 wearing each other participant's (P) custom-fitted earpieces, for concha (C), canal-front (F) and canal-back (B).}
\label{tab:imposter_impedances}
\end{table}

@@

Since every participant has one classifier per task (for which that task is the passthought), we are able to make 200 spoofed attempts with the correct passthought on each of P6's classifiers. We find zero successful spoof attempts for tasks with a chosen secret (e.g., \textit{song} or \textit{face}). In addition, we also do not find any successful spoof attacks for tasks with no chosen secret (e.g., \textit{breathe}). In fact, in all 1,800 spoof attempts (200 attempts for each of the nine classifiers), we do not find a single successful attack on any of P6's classifiers.

@@

Our findings demonstrate the apparent feasibility of a passthoughts system consisting of a single earpiece with three electrodes and a reference, all in or on the left ear. FARs and FRRs are very low across all participants and tasks, with FARs overall lower than FRRs, a desirable pattern in terms of authenticating access to potentially sensitive information. Participants' best-performing tasks or passthoughts typically see virtually no errors in our training. Passthoughts also appear to be quite memorable given our two week recall follow-up and some were rated highly repeatable and engaging. Furthermore, no spoofed attacks were successful in our cursory analyses. These results are a marked improvement upon the only other in-ear EEG passthoughts work we are aware of \cite{curran2016passthoughts}, from 80\% accuracy to 99.82\%. This improvement is likely due to the custom-fit quality of the earpieces, the use of three electrodes in the ear as opposed to one, and possibly the XGBoost classifier instead of a generic support vector machine.

@@

Our multiple classifier training strategies establish the presence of both knowledge and inherence factors as schemes isolating each still achieve low FARs. Interestingly, by training classifiers to reject both wrong participants and wrong tasks from the correct participant, thus incorporating both the knowledge and inherence factors, we see an increase in FRR suggesting multi-factor authentication's added security may come at a small cost to convenience. As preliminary evidence of a possession factor in the current prototypes, we had P1 use their best task (listen) with P6's earpieces but tested on P1's classifier, with a resultant FAR of 0. The custom-fit earpieces could alo easily include a hardware keypair to sign authentication attempts presenting a more explicit possession factor.

The powerful interactions between inherence and knowledge emerged in our spoofing attacks. Although our target participant documented their chosen passthought, the spoofers found ambiguity in how these passthoughts could be expressed. For the \textit{face} task, the spoofers did not know the precise face the original participant had chosen. For the \textit{song} tasks, though the song was known, the spoofers did not know what part of the song the original participant had imagined, or how it was imagined. This experience sheds light on passthoughts' highly individual nature and suggests there may be intrinsic difficulty in spoofing attempts.

Contrary to our expectations, relatively low impedances were still achievable by imposters using another participants' custom-fit earpieces, despite the uniqueness of ear canal shapes between individuals \cite{Akkermans2005}. This may have been due to our use of conductive gel and warrants further exploration. This discovery is not detrimental, however, as the primary inherence factor of this system is drawn from the uniqueness of EEG and not ear canal shape.

@@

The real-world implications of this study are limited by the small, relatively homogeneous sample of participants. In the case of this initial evaluation however, the homogeneity of our participant pool can be seen as a strength of the reported results given that system is meant to distinguish between individuals. In order to establish the validity of this system for widespread real-world use we feel it is necessary to expand the size and diversity of participants in future studies. This sample size is comparable to that of other EEG passthoughts work \cite{Ashby2011, Marcel2007a, Poulos2002, Chuang2013b, curran2016passthoughts} and other custom-fit in-ear EEG research \cite{Kidmose2013a, Mikkelsen2015}. The fitting and manufacturing of custom-fit earpieces was the main limitation to increasing our sample size. This may very well pose a limitation in proliferation and adoption of such a technology as well, although recently there have been developments in at-home kits for creating one's own custom-fitted earpieces \cite{voix2015settable} that could help overcome this barrier.

@@

One primary question is how our passthoughts system performance will change with a greater number of users and with more diverse data. Our system specifically trains on negative examples of incorrect users; we do not yet know how this approach will scale. At the same time, we must investigate the stability of EEG readings for a passthought over time to establish long-term usability. We must also collect EEG data from the variety of different user states: ambulatory settings, during physical exertion or exercise, under the influence of caffeine or alcohol, etc.

>>>>

\title{Exploring the Feasability of\\ Multifactor Authentication with EarEEG}

@@

\begin{abstract}
Multifactor authentication presents a robust method to secure our private information, but typically requires multiple actions by the user resulting in a high cost to usability and limiting adoption. A usable system should also be unobtrusive and inconspicuous. We present and discuss a system with the potential to engage all three factors of authentication (inherence, knowledge, and possession) in a single step using an earpiece that implements brain-based authentication using electroencephalography (EEG). We demonstrate its potential by collecting EEG data using manufactured custom-fit earpieces with embedded electrodes and testing a variety of authentication scenarios. Across all participants’ best-performing “passthoughts”, we are able to achieve 0\% false acceptance and 0.36\% false rejection rates, an overall accuracy of 99.82\%, using one earpiece with three electrodes. Furthermore, we find no successful attempts simulating impersonation attacks. We also report perspectives from our participants and discuss directions for future work to explore earEEG for the masses. Our results indicate that a relatively inexpensive system using a single electrode-laden earpiece could provide a discreet, convenient, and robust method for one-step, three-factor authentication.

@@

We find that we can achieve very low false acceptance rates (FAR) and false rejection rates (FRR) with a single, three-electrode earpiece. Interestingly, we found that performance improves in the ear compared to a single electrode on forehead above the left eye, a site typically used for collecting EEG data. We also find that passthoughts could not be spoofed by imposters, even when the imposters had the user's earpiece and knew their chosen passthoughts. These results improve upon previous work on earEEG-based passthoughts systems by testing custom-made earpieces with multiple embedded electrodes which achieve above 99\% performance, an approximately 20\% improvement \cite{curran2016passthoughts}. We also detail important aspects of the system including measurements of the electrodes' electrical impedances and the results of various classifier training and testing schema to probe each authenetication factor. Finally we discuss interesting patterns in our promising results, and future work to investigate the robustness of passthoughts in day-to-day life.

@@

These results thus far establish good performance in our default training strategy, in which we count as negative examples recordings from the wrong participant performing any task. For comparison, we report three other analyses with differing negative examples which serve to isolate and test the inherence and knowledge factors: the correct task recorded from the wrong participant (relies on inherence only), the wrong task recorded from the correct participant (relies on knowledge only), and a combination of these two. Positive examples were always the correct participant performing the correct task.

@@

Overall, our default training strategy which engages both knowledge and inherence factors achieves the lowest FAR (Table \ref{tab:compare}). The FAR in the inherence-only scenario is ten times higher, and the knowledge-only FAR is one hundred times higher, though for all scenarios FAR is less than 1\%. However, FRR is \textit{lower} in the inherence-only training strategy than the default. The highest FRR is in the combined negative example case, though FAR remains low.

Our "leave one out" analysis with participants' best tasks maintained 0 FAR across all participant combinations.

@@

The first aspect of this scenario we tested was the ability of an imposter to wear an earpiece acquired from someone else and achieve viable impedance values for EEG collection based on the fit of the pieces in their ears. P1 tried on each of the other participants' customized earpieces, which were able to at least physically fit in P1's ears. The impedances of P1 wearing other participants' earpieces were then recorded and are listed in Table \ref{tab:p1_imposter_impedances} below. While some are a better fit than others, overall they are higher (worse) than those achieved by the pieces' intended owners themselves.

\begin{table}[h]
\begin{center}
\begin{tabular}{lrrrrrr}
& \multicolumn{6}{c}{Impedance [k\(\Omega\)]} \\
\cline{2-7}
& \multicolumn{3}{|c|}{\textbf{Left ear}} & \multicolumn{3}{c|}{\textbf{Right ear}} \\
\textbf{P} & \textbf{C} & \textbf{F} & \textbf{B} & \textbf{C} & \textbf{F} & \textbf{B} \\
\hline
2 & 34.1 & 10.2 & 12.8 & 27.8 & 16.0 & 16.3\\
3 & 21.1 & 20.9 & 19.0 & 13.5 & 11.3 & 19.5\\
4 & 14.1 & 11.9 & 9.7 & 11.0 & 11.1 & 13.3\\
5 & 17.2 & 21.9 & 10.3 & 32.6 & 12.5 & 11.6\\
6 & 18.7 & 10.0 & 8.4 & 14.8 & 11.5 & 8.9\\
7 & 91.5 & \textgreater1000 & 21.5 & 33.5 & 26.4 & 31.0\\
\end{tabular}
\end{center}
\caption{Electrical impedances with P1 wearing each other participant's (P) custom-fitted earpieces, for concha (C), canal-front (F) and canal-back (B).}
\label{tab:p1_imposter_impedances}
\end{table}

@@

Since every participant has one classifier per task (for which that task is the passthought), we are able to make 200 spoofed attempts with the correct passthought on each of P6's classifiers. We find zero successful spoof attempts for tasks with a chosen secret (e.g., \textit{song} or \textit{face}), see Table \ref{tab:imposter_impedances}. In addition, we also do not find any successful spoof attacks for tasks with no chosen secret (e.g., \textit{breathe}). In fact, in all 1,800 spoof attempts (200 attempts for each of the nine classifiers), we do not find a single successful attack on any of P6's classifiers.

@@

Our findings demonstrate the apparent feasibility of a passthoughts system consisting of a single earpiece with three electrodes and a reference, all in or on the left ear. FARs and FRRs are very low across all participants and tasks, with FARs overall lower than FRRs, a desirable pattern in terms of authenticating access to potentially sensitive information. Participants' best-performing tasks or passthoughts typically see virtually no errors in our training. From our various training/testing schema it emerged that the inherence factor performs better on its own compared to the knowledge factor, but the combination of the two achieves the lowest FAR indicating measurable benefit of multiple factors. Passthoughts also appear to be quite memorable given our two week recall follow-up and some were rated highly repeatable and engaging. Furthermore, no spoofed attacks were successful in our cursory analyses. These results are a marked improvement upon the only other in-ear EEG passthoughts work we are aware of \cite{curran2016passthoughts}, from 80\% accuracy to 99.82\%. This improvement is likely due to the custom-fit quality of the earpieces, the use of three electrodes in the ear as opposed to one, and possibly the XGBoost classifier instead of a generic support vector machine.

@@

Our multiple classifier training strategies establish the presence of both knowledge and inherence factors as schemes isolating each still achieve low FARs. Interestingly, by training classifiers to reject both wrong participants and wrong tasks from the correct participant, thus incorporating both the knowledge and inherence factors, we see an increase in FRR suggesting multi-factor authentication's added security may come at a small cost to convenience. As preliminary evidence of the potential \textit{secondary} inherence factor of ear shape in the current prototypes, we tested P1 using their best task (listen) with P6's earpieces but tested on P1's classifier, with a resultant FAR of 0, suggesting the protective value of the custom-fit earpieces likely due to differing electrical impedances from one user to another. To add a possession factor, these custom-fit earpieces could also easily include a hardware keypair to sign authentication attempts, functioning similarly to the way a key fob can be used to unlock a car, but compared to this single-factor system the demonstrated inherence and knowledge factors would provide important extra layers of security.

Contrary to our expectations, relatively low impedances were still achievable by imposters using another participants' custom-fit earpieces, despite the uniqueness of ear canal shapes between individuals \cite{Akkermans2005}. This may have been due to our use of conductive gel and warrants further exploration. This discovery is not detrimental to the system, however, as the primary inherence factor of this system is drawn from the uniqueness of individuals' EEG patterns and not ear canal shape.

The powerful interactions between inherence and knowledge emerged in our spoofing attacks. Although our target participant documented their chosen passthought, the spoofers found ambiguity in how these passthoughts could be expressed. For the \textit{face} task, the spoofers did not know the precise face the original participant had chosen. For the \textit{song} tasks, though the song was known, the spoofers did not know what part of the song the original participant had imagined, or how it was imagined. This experience sheds light on passthoughts' highly individual nature and suggests there may be intrinsic difficulty in spoofing attempts. Future work should examine this effect more explicitly to elucidate the effect of knowledge task specificty on defense against imposters.

@@

The real-world implications of this study are limited by the controlled laboratory setting in which it was conducted as well as the small, relatively homogeneous sample of participants. In the case of this initial evaluation however, the homogeneity of our participant pool can be seen as a strength of the reported results given that system is meant to distinguish between individuals. In order to establish the validity of this system for widespread real-world use we feel it is necessary to expand the size and diversity of participants in future studies, encompassing users for whom the system would be particularly attractive such as those with extreme security needs and/or persons with disabilities which prevent them from performing other authentication methods. This sample size is comparable to that of other EEG passthoughts work \cite{Ashby2011, Marcel2007a, Poulos2002, Chuang2013b, curran2016passthoughts} and other custom-fit in-ear EEG research \cite{Kidmose2013a, Mikkelsen2015}. The fitting and manufacturing of custom-fit earpieces was the main limitation to increasing our sample size. This may very well pose a limitation in proliferation and adoption of such a technology as well, although recently there have been developments in at-home kits for creating one's own custom-fitted earpieces \cite{voix2015settable} that could help overcome this barrier.

@@

One primary question is how our passthoughts system performance will change with a greater number of users and with more diverse data. Our system specifically trains on negative examples of incorrect users; we do not yet know how this approach will scale. At the same time, we must investigate the stability of EEG readings for a passthought over time to establish long-term usability and determine when a classifier retraining period is necessary. Recent work in scalp-based BCI research has demonstrated individuals' EEG permanence over one to six months \cite{Maiorana2016, Armstrong2015} or even over one year \cite{Ruiz2017} between sessions, though the permanence of passthougts as described here and in-ear EEG collection remains unexplored to our knowledge. Future work must also examine EEG data collected from a variety of different user states: ambulatory settings, during physical exertion or exercise, under the influence of caffeine or alcohol, etc.
